**Reference-based OTU clustering for machine learning classification**

### 2022-06-15 submission/manuscript.Rmd
full manuscript draft

---
title: "**TITLE**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
header-includes:
  - \usepackage{helvet}
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{20mm}

@@

\vspace{20mm}

Courtney R. Armour${^1}$, William L. Close$^{1,*}$, Begüm D. Topçuoğlu$^{1,\#}$, Patrick D. Schloss $^{1,\dagger}$

@@

${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor MI.

@@

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

\vspace{20mm}

@@

@@

@@

## Introduction (~250 words)  
- Gut microbiome community composition has proven useful as a resource for machine learning prediction of various diseases {examples}. 
- Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. 
- Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
- However OTU clustering depends on the data in the dataset and the addition of new data may change the overall OTU clusters. 
- The unstable nature of OTU clustering complicates machine learning. When building a classification model, changes to the OTU clusters means you have to re-create the model. This can change the underlying model and be time consuming and resource intensive. 
- The ability to integrate new data into an existing model could allow for deployment of a single model that new data can be continually added to and predicted on.
- Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. 
- While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
- Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.

## Results (~700 words)
- Utilize public dataset with normal and SRN samples to compare prediction between OptiFit and OptiClust
- randomly split data into 80% training 20% test sets 100 times
- Processed the data with both algorithms - Figure 1
  - 1. Used traditional OptiClust method to cluster all data, then split into training and test set
  - 2. Used OptiClust on the 80% training set, then used OptiFit to fit the remaining 20%
- OptiFit produces similar quality OTU clusters based on MCC (supplement?)
- Used mikropml to train a model on the 80% training set for each data split, then predicted dx on the 20% test set
- Training performance almost identical (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Performance on the test set comparable (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Maybe expand on where optifit did better/worse?

## Discussion/Conclusions (~250 words)
- OptiFit works!
- future questions:
  - how much reference do you need?
  - does it work well for other situations?
    - what if we used a new dataset for test set instead of a subset of the full dataset?
    - other diseases?

## Materials and Methods
- 16S rRNA amplicon sequence data from 490 subjects {baxter}
  - 261 controls
  - 229 SRN
- created 100 random splits of the data (80% training, 20% test)
- preprocessed data with mothur v1.45
- two pathways  
  -opticlust:  
    - clustered all data together
    - split the shared file based on the random splits

  -optifit:  
    - split the data
    - opticlust on the 80% training set
    - optifit to fit the remaining 20% to the training set OTUs
- ML with mikropml package (version)
  - preprocessed training set and applied preprocessing to test set
    - correlated collapsed, removed nzv

## Acknowledgements
- funding

**Figure 1. Workflow** description.

**Figure 2. Model Performance.** **A)** Mean AUC **B)** Averaged ROC curves

@@

>>>>

---
title: "**Reference-based OTU clustering for machine learning classification**"
output:
  rmarkdown::pdf_document:
    keep_tex: yes
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: mbio.csl
link-citations: true
---

```{r settings, eval=TRUE, echo=FALSE, cache=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

format_decimal <- function(number,digits){
  return(round(number,digits=digits))
}

```

Other Title Options:\
A solution to the problem of inconsistent OTU clustering for machine
learning classification

Optimized reference-based OTU clustering for machine learning
classification

@@

@@

Courtney R. Armour${^1}$, Kelly L. Sovacool${^2}$, William L.
Close$^{1,*}$, Begüm D. Topçuoğlu$^{1,\#}$, Jenna Wiens${^3}$, Patrick
D. Schloss $^{1,\dagger}$

${^1}$ Department of Microbiology and Immunology, University of
Michigan, Ann Arbor, Michigan, USA

${^2}$ Department of Computational Medicine and Bioinformatics,
University of Michigan, Ann Arbor, Michigan, USA

${^3}$ Department of Electrical Engineering and Computer Science,
University of Michigan, Ann Arbor, Michigan, USA

@@

$\dagger$ To whom correspondence should be addressed:
[pschloss\@umich.edu](mailto:pschloss@umich.edu)

\vspace{10mm}

@@

Machine learning classification of disease based on the gut microbiome
often relies on operational taxonomic units (OTUs) to quantify microbial
composition. The standard methodology for quantifying taxonomic
composition is clustering 16S rRNA amplicon sequences *de novo* into
OTUs and using their abundances to train a classification model.
However, OTU clusters depend on the sequences in the data set and
therefore can change if new data is added. This inconsistency
complicates machine learning classification because in order to use the
model to classify additional samples, the OTUs have to be re-clustered
to include the new sequences and the model retrained with the new OTU
clusters. A new algorithm in mothur, called OptiFit, addresses this
issue by clustering new sequences into the existing OTU clusters. While
OptiFit is proven to produce high quality OTU clusters, it is unclear
whether this method for clustering new sequence data into existing OTUs
will have any impact on machine learning classification using the OTU
abundances. To evaluate the potential for use in machine learning
classification, we used OptiFit to cluster additional data into existing
OTU clusters and quantified model performance in classifying a data set
containing samples from controls and people with advanced adenoma and
carcinoma (SRN). We compared the model performance using OptiFit to the
standard procedure of clustering all the data together. We found that
OptiFit can be used to fit new sequence data to existing OTUs and
perform equally in classifying SRNs. Moving forward, when OTUs are used
in machine learning classification problems, OptiFit can be used to
avoid the need to re-cluster sequences into OTUs when classifying
additional samples that were not part of the data set used to build the
model.

@@

There are instances when OTU abundance data is optimal for machine
learning prediction of disease based on microbiome composition. The
current methodology for OTU clustering complicates machine learning
since *de novo* OTU clusters can change depending on the data in the
data set. Attempting to classify additional samples that were not part
of the original model build requires re-clustering OTUs and re-training
the model. OptiFit provides an elegant solution to the problem of OTU
inconsistency by enabling new data to be fit to the existing OTUs while
maintaining machine learning classification performance. The use of
OptiFit will enable a single model based on OTU abundance data to be
used in classifying additional samples that were not part of the
original OTU clustering.

@@

Gut microbiome community composition is useful as a resource for machine
learning prediction of diseases, including colorectal cancer
[@baxter2016; @zackular2014]. Amplicon sequencing of the 16S rRNA gene
is a reliable tool for assessing the taxonomic composition of microbial
communities. Analysis of 16S rRNA amplicon sequence (16S) data generally
relies on clustering of sequences based on similarity into operational
taxonomic units (OTUs). The process of OTU clustering can either be
reference-based or *de novo*. The quality of OTUs generated with
reference-based clustering is dependent on the quality of the reference
used, while OTUs generated with *de novo* clustering are optimized to
the data set @westcott2015. While *de novo* clustering can produce
high-quality OTU clusters where sequences are accurately grouped based
on similarity thresholds, the resulting OTU clusters depend on the data
in the data set and the addition of new data could change the overall
OTU clusters. The inconsistent nature of OTU clustering complicates
deployment of machine learning models since integration of additional
data requires re-clustering all the data and re-training of the model.
The ability to integrate new data into an existing model without
re-clustering and re-training could allow for deployment of a single
model that new data can be continually added to. Recently Sovacool *et
al* introduced OptiFit: a method for fitting new data into existing OTU
clusters @sovacool2022. While OptiFit is proven to effectively fit new
sequence data to existing OTU clusters, it is unknown if the use of
OptiFit will have an impact on machine learning classification. Here we
test the ability of OptiFit to cluster new sequence data into existing
OTU clusters for the purpose of machine learning classification of
disease based on gut microbiome composition.

When using OTU abundances for machine learning classification, the
current methodology is to *de novo* cluster all of the sequence data
into OTUs with the OptiClust algorithm in mothur @westcott2017. The
resulting abundance data is then split into training and testing sets,
where the training set is used to tune hyperparameters and ultimately
train the model. The testing set is then classified with the model and
the performance of the model can be quantified (Figure 1A). However,
with this methodology we would have to re-generate the OTU clusters and
re-train the model if we wanted to classify additional samples. The
OptiFit algorithm @sovacool2022 addresses this problem by enabling new
sequences to be clustered into existing OTUs. The OptiFit workflow is
similar to the OptiClust workflow where the data is clustered into OTUs
and used to tune hyperparameters and ultimately train the model. Then,
we can use OptiFit to fit sequence data of samples not part of the
original data set into the existing OTUs and use the same model to
classify the samples (Figure 1B). To test how the model performance
compares between these two methodologies, we used a publicly available
data set of 16S sequences from stool samples of healthy subjects as well
as subjects with SRN consisting of advanced adenoma and carcinoma
@baxter2016. The data set was randomly split into an 80% train set and
20% test set. For the standard OptiClust workflow, the training and test
sets were *de novo* clustered together into OTUs then the resulting
abundance table was split into the training and testing set. For the
OptiFit workflow, the train set was clustered *de novo* into OTUs and
the remaining test set was fit to the OTU clusters using the OptiFit
algorithm. For both workflows, the abundance table of the train set was
used to tune hyperparameters and train a random forest model to classify
SRN. The test set was classified as either control or SRN using the
trained models.To account for variation depending on the split of the
data, the data set was randomly split 100 times and the process repeated
for each of the 100 data splits. By comparing the model performance of
classifying the samples in the test data set between the OptiFit and
OptiClust algorithms, we can quantify the impact of using OptiFit on
model classification performance.

```{r mcc,include=F}
merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")

opticlust_mcc <- merged_mcc %>% 
  filter(algorithm == "opticlust") %>% 
  pull(mcc)

optifit_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc))
```

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score is quantified by examining all pairs of sequences and assessing
whether they belong together in an OTU or not based on sequence
similarity @westcott2017. MCC scores range between negative one and one.
A score of negative one means none of the sequences in an OTU are within
the similarity threshold and any sequences within the similarity
threshold are not in an OTU together. An MCC score of zero essentially
means the sequences are randomly clustered. An MCC score of 1 means all
sequences in an OTU are within the similarity threshold and all sequence
pairs within the similarity threshold are in the same OTU. To ensure
that OptiFit is appropriately integrating new sequence data into the
existing OTUs, we would expect that the MCC scores produced by the
OptiClust and OptiFit workflows are similar. Since the data is only
clustered once in the OptiClust workflow there is only one MCC score
while the OptiFit workflow produces an MCC score for the OTU clusters
from each data split. Overall the MCC scores were similar between
OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average
MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit
performs as well as OptiClust when integrating new sequences into the
existing OTUs.

```{r performance,include=F}
performance <- read_csv("../data/learning/summary/merged_performance.csv",
                    col_types = cols(algorithm=col_character(),
                                     method=col_character(),
                                     split=col_character(),
                                     .default=col_double()))

performance_summary <- performance %>% 
  group_by(algorithm) %>% 
  summarise(median_cv_AUC = round(median(cv_metric_AUC),digits=3),
            mean_cv_AUC = round(mean(cv_metric_AUC),digits=3),
            median_AUC = round(median(AUC),digits=3),
            mean_AUC = round(mean(AUC),digits=3))

pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
                                                                   .default = col_character())) %>% 
  select(metric,p_value)
```

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we next examined the model performance for
classifying samples in the test data set as control or SRN. To quantify
model performance we used the taxonomic abundances of the training data
from the OptiClust and OptiFit workflows to train a model to predict
SRNs. Using the predicted and actual diagnosis classification, we
calculated the area under the receiver operating characteristic curve
(AUROC) for each data split to quantify model performance. During
cross-validation (CV) training, the model performance was equivalent
between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
OptiClust mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
OptiFit mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
Figure 2A). The trained model was then deployed to classify the samples
of the test data as control or SRN. The performance on the test data was
equivalent between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
OptiClust mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
OptiFit mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
Figure 2B,C) indicating that new data can be fit to existing OTU
clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification of disease based
on microbiome abundance. A potential problem to using OptiFit for
machine learning prediction is that any sequences in the new data that
do not map to the existing OTU clusters will be discarded resulting in a
possible loss of information. However, we demonstrated that OptiFit can
be used to fit new sequence data into existing OTU clusters and perform
equally in predicting SRN compared to clustering all of the sequence
data together. The ability to integrate new data into existing OTUs
enables the deployment of a single machine learning model based on
microbiome composition that new data can be classified with. These
results are based on a single data set and disease. Further analysis is
needed to determine the data size necessary to build a robust model
capable of classifying diverse data sets. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN.

## Materials and Methods

***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human
stool samples was downloaded from NCBI Sequence Read Archive (accession
no. SRP062005) @baxter2016.; @edgar2011 This data set contains stool
samples from a total of N subjects, however after preprocessing to
screen for sequence quality and subsample to 10,000 reads per sample,
490 samples remained. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as "normal" while samples from subjects identified as
advanced adenoma or carcinoma were categorized as "screen relevant
neoplasia" (SRN). The resulting data set consisted of 261 normal samples
and 229 SRN samples.

```{r n_test_train,include=F}
n_test_train <- read_csv("../results/tables/splitTogetherFrequency.csv") %>% 
  mutate(n_test = rowSums(across(starts_with("split_")))) %>% 
  select(Group,n_test) %>% 
  mutate(n_train = 100 - n_test) %>% 
  summarize(avg_test = mean(n_test),
            sd_test = sd(n_test),
            avg_train = mean(n_train),
            sd_train = sd(n_train))
```

***Data Processing.*** The full dataset was pre-processed with mothur
(v1.45) @schloss2009 using the SILVA reference database (v132)
@quast2013 to join forward and reverse reads, merge duplicate reads,
align to the reference, pre-cluster, remove chimeras with UCHIME
@edgar2011, assign taxonomy, and remove non-bacterial reads following
the Schloss Lab MiSeq standard operating procedure described on the
mothur website (<https://mothur.org/wiki/miseq_sop/>). 100 splits of the
490 samples were generated where 80% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20% (98 samples)
were assigned to the test set. Using 100 splits of the data accounts for
the variation that may be observed depending on the samples that are in
the training or test sets. Each samples was in the training set an
average of `r n_test_train %>% pull(avg_train)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
the test set an average of `r n_test_train %>% pull(avg_test)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm @westcott2017. In this pathway,
all of the data was clustered together with OptiClust to generate OTUs
and the resulting abundance tables were split into the training and
testing sets. In the second workflow, the pre-processed data was split
into the training and testing sets. The training set was clustered into
OTUs, then the test set was fit to the OTUs of the training set using
the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
method open so that any sequences that didn't map to the existing OTU
clusters would form new OTUs. Any OTUs that were not in the training set
were removed prior to machine learning. For both pathways, the shared
files were sub-sampled to 10,000 reads per sample.

***Machine Learning.*** Machine learning using Random Forest was
conducted with the R package mikrompl (v XXXX) @topçuoglu2021 to predict
the diagnosis (SRN or normal) for the samples in the test set for each
data split. The training set was preprocessed to normalize values
(scale/center), collapse correlated features, and remove features with
zero-variance. The preprocessing from the training set was then applied
to the test set. P values comparing model performance were calculated as
previously described {}. The averaged ROC curves were plotted by taking
the average and standard deviation of the sensitivity at each
specificity value.

***Code Availability.*** The analysis workflow was implemented in
Snakemake @koster2012 . Scripts for analysis were written in R @R2020
and GNU bash @GNUbash. The software used includes mothur v1.47.0
@schloss2009, RStudio @RStudio2019, the Tidyverse metapackage
@wickham2019, R Markdown @xie_r_2018, the SRA toolkit
@noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016. The
complete workflow and supporting files required to reproduce this study
are available at:
<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021>

## Acknowledgements

(funding)

@@

```{=tex}
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
```
\noindent

::: {#refs}
:::

```{=tex}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
```
\newpage

![](../exploratory/figures/figure1_a.pdf)

![](../exploratory/figures/figure1_b.pdf)

**Figure 1: Workflows.** **A)** OptiClust workflow: The full data set
was clustered into OTUs using the OptiClust algorithm in mothur. The
data was then split into two sets where 80% of the samples were assigned
to the training set and 20% to the testing set. The training set was
preprocessed with mikropml to normalize values (scale/center), collapse
correlated features, and remove features with zero-variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the standard OptiClust
workflow. **B)** OptiFit workflow: The data set was first split into two
sets where 80% of the samples were assigned to the training set and 20%
to the testing set. The training set was then clustered into OTUs using
the OptiClust algorithm in mothur. The resulting abundance data was
preprocessed with mikropml to normalize values (scale/center), collapse
correlated features, and remove features with zero-variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The OptiFit algorithm in mothur was used to
cluster the left out testing data set using the OTUs of the training set
as a reference. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the new OptiFit
workflow.

![](../exploratory/figures/figure2.pdf)

**Figure 2: Model Performance.** **A)** Area under the receiver
operating characteristic (AUROC) curve during cross-validation for the
OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. Mean AUROC is
printed to the right of the points. **B)** AUROC on the test data for
the OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. Mean AUROC is
printed to the right of the points. **C)** Receiver operating
characteristic (ROC)Averaged ROC curves
